% !TeX root = main.tex

\section{Session Types}
%\label{sec:preliminaries}
\label{sec:session_types}


This section introduces session types through
a discussion of the need for them in
software engineering.
This section takes the approach to identify and define
a clear and distinctive terminology
for transferring the basic ideas of session types to a wider audience.
It also presents the Scribble~\cite{scribble} protocol
description language, which is a formal language for describing
protocols using the principles of session types.

\subsection{Session Types and Software Engineering}%: Notions and Terminology}
\label{sec:sessions_software}

\input{figures/fig-sequence_diagram}

Communication \emph{protocols} in software engineering
are usually described and communicated among humans
by using the most intuitive graphical form of sequence diagrams
%(\figref{seqchart}).
Sequence diagrams are used to depict a communication flow as an ordered
sequence of \emph{message passing} among communicating
participants that represent particular \emph{roles} in a communication
protocol.

\figref{seqchart} illustrates the bookstore protocol, which is a use-case
in the ABCD online repository~\cite{usecase_repository}.
The bookstore protocol specifies an interaction between two online Buyer roles
that invoke a process of buying a book from an online Seller role, while
sharing the expense. Intuitively the communication (horizontal lines) among
roles (vertical lines) proceeds downwards.
%In the sequence diagram, each role is represented by a vertical line and
%communication messages are represented by horizontal arrows
%indicating the direction of communication flow.
%Intuitively, the communication progress proceeds downwards.
Thus the protocol's behaviour is represented visually within the diagram.
However, sequence diagrams are not considered as formal tools because
protocols described in this way cannot be verified as being complete or unambiguous.
%
In industrial practice, communication protocols for
distributed software are usually described in natural language and
discussed with the help of sequence diagrams.
%After that and with the help of sequence charts,
%the protocols are implemented between the communication components
%of the software.
%within the distributed software
%with specified communication protocol.
Each software module that participates in the
protocol implements
its \emph{local} software role within the protocol.
%Usually the process of implementing the local specifications
%may follow different possible solutions that are not directly
%visible or are even omitted from the sequence diagram, due
%to the complexity they will introduce. \sg{I'm not sure what this means.}
%\dk{what is hidden? - overall structure of the protocol, logic on choices}
A limitation of sequence diagrams is the inability to deal with complex
communication protocols, such as the case where one role
can make a choice that affects the subsequent communication behaviour
of the entire protocol.

Communication complexity can be managed by
introducing \emph{structure} within the communication protocol.
Session types describe communication in a structured and rigorous fashion,
and aim to define concepts for structured communication
that should enable better control of communication complexity.

%describing communication protocols
Typically, a type is broadly understood as a concept used
to describe a class of data. Structured types describe
data-structures. When types structure input/output requirements
they can describe functionality. Combination of functionality
and data structure give rise to object-oriented concepts.
A session type extends
the notion of a type to capture the communication behaviour.% in concurrent systems.
{\em A session type is a type structure that describes communication
	behaviour in terms of a series of send and receive
	interactions between a set of communicating entities.} 

%Session types can be seen as types that describe behaviour.
%Typically, a type is broadly understood
%as a meta-information concept, that is used to describe a class of data.
%Structured types are used to describe data structures.
%When types are structured as input/output requirements they
%are used to describe functionality.
%The combination of functionality and data structures give rise
%to the \emph{object-oriented} concepts of classes and interfaces. % of objects.
%%
%A session type extends %the above approaches of
%the notion of a type to capture the communication behaviour in concurrent systems.
%% definition of session types
%{\em A session type is a type structure that describes communication
%	behaviour in terms of a series of send and receive %message passing
%	interactions between a set of communicating entities.}  %, and it is used
%%	to characterise data structures that are used to implement communication.} \sg{I'm not sure what we mean by characterising data structures. Is it just that the data type of each message is defined?}

The formal nature of session types enables rigorous
reasoning about the communication properties of a system.
Session types %particularly
focus on the %following
properties:
%
\begin{enumerate}[label=$\bullet$]
	\item	{\em Communication Safety}: Every execution state of the system has {\em safe} communication behaviour:
	%
	\begin{itemize}
		\item	{\em Communication interaction match}:
				Every send
				%(resp., select)
				operation has a corresponding receive
				%(resp., branch)
				operation.
		\item	{\em Message type match}:
				The type of every message being sent matches the type of the message expected to be received.
		\item	{\em Deadlock-freedom}:
				%Consequently,
				Every message sent will eventually be received.
	\end{itemize}

	\item	{\em Communication Progress}: If the protocol has not yet terminated, then there is a pair of components that can communicate safely, leading to a system state that again satisfies the safety and progress properties.
%Every state of the system either represents a terminated protocol, or
%			will {\em progress} to a state with safe communication behaviour. \sg{Maybe this can be clearer?}
\end{enumerate}
%
%In addition, the structured nature of session types enables
%a methodology for integrating session types in software development.

%\dk{make clear that it is a syntactic/static technique - in contrast to model checking}

Furthermore, the Scribble tool-chain~\cite{scribble} is a platform
that intends to serve as a core element in the process
of integrating the practical aspects of session types
in software engineering. The design of Scribble
draws directly from the principles and properties of session types,
as identified through rigorous study. The rest of this
section introduces the main terminology for session types
and, whenever appropriate, it presents these notions
using the Scribble tool-chain.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Session Types as Protocol
%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\subsection{Sessions Types as Protocols}%: The Scribble Protocol Description Language}
\label{sec:sessions_scribble}

%The Scribble tool-chain~\cite{scribble} is a tool that indents
%to serve as a core element in the process of integrating
%the practical aspects of session types in software engineering.
%The basic module of the Scribble
%tool-chain is the Scribble protocol description language,
%which is a syntax that expresses session type specifications.
%The design of the Scribble language draws directly from
%the principles of session types.

Practically, a session type can be seen as a formal
{\em specification} of a communication {\em protocol} among communicating {\em roles}.
The syntax for session types is defined on pairs of operations:
(i) the {\em send} and {\em receive} operation give rise to
{\em message passing}; and
(ii) the {\em select} and {\em branch} operators enable the
{\em choice} interaction.

Message passing is used for sending data from one role to another.
The data that are passed are described in terms of their type. Session
types can also support the case of {\em session delegation}
where a session-typed value, i.e.~a structure that implements communication,
is passed as a message.
Choice requires a role to use a value called a {\em label}
to select a communication behaviour among
a set of behaviours, called {\em branch}, offered by another role.
The choice interaction enables formal specifications that can
cover a more complex structure of communication and describe all possible
cases that may happen, in contrast to the limiting structure of
a sequence diagram.
%The syntax for session types is defined on {\em send} and {\em receive},
%and {\em select} and {\em branch} operators.
%Sending and receiving data gives rise to the notion of {\em message passing}.
%In a session type messages are described via their types.
%The case where a session typed value is passed as a message is called {\em session delegation}.
%The interaction of select and branch defines the {\em choice} interaction.
%Choice requires for a role to use a value called {\em label}
%to select a communication behaviour out of a set of behaviours, called {\em branch},
%offered by another role.
%\textbf{Important is to note that branch and choice elements within the formal specification force
%the user to think among all possible cases that may happen within an session type.
%Comparing to sequence charts which are not formal and thus more error prone.}

Session types assume a {\em global} protocol that describes the
communication interaction of all the roles inside a concurrent system.
A global type enables a view of a system as a whole, instead of the
approach that a concurrent system is a set of communicating modules.
The perspective of a global protocol through a single role
is called {\em local} protocol. The local protocol describes
the communication interaction of a single role with all other roles
in the system. Local protocols can also be seen as syntactic
representations of state machines, where each state machine edge
depicts a communication operation.

The relation between a global protocol and the
local protocols of its roles, is expressed
through the {\em projection} procedure;
% called {\em projection}.
%Projection to a role
%is a procedure that can be done
%can be done automatically;
given the global protocol and a role, projection
returns as a local protocol only the
interactions of the global protocol that are concerned with
the role.
%
The reverse procedure to projection is called {\em synthesis},
where a set of local protocols are composed %together
in a global protocol.


\subsection{The Scribble Protocol Description Language}
\label{sec:scribble}
The notions explained above are supported by the
Scribble protocol description language,
which is a syntax used to express session type specifications.
As an example,
consider part of the session type protocol for the book-store
use-case as in \figref{scribble_bs}.


\input{figures/fig-scribble_bookstore}

Line 1 of the upper protocol in \figref{scribble_bs} shows the
Scribble definition of the global protocol.
The communication takes place
among the roles \BuyerOne, \BuyerTwo and \Seller. Line 2 demonstrates
the syntax of a message passing interaction; a message
of type \Code{title} labelled with label \Code{book} is sent \lstinline|from|
role \BuyerOne \lstinline|to| role \Seller. Scribble, defines a message 
as a list of types annotated with a label.

In line 5 of the protocol
%their properties
there is a choice interaction,
where role \BuyerTwo makes an internal decision and selects
from two possibilities, that are
expressed with the labels \Code{agree} in line 5 or \Code{quit} in line 6.
In  both cases roles \BuyerOne and \Seller respectively offer
alternative branches. The choice interaction requires that
Role \BuyerTwo informs roles \BuyerOne and role \Seller. All
three roles will then synchronise on the same choice.

Scribble also provides a looping construct, which is needed for more realistic protocols but is not used in this simple example.

%\dk{describe the recursion construct?}

Line 1 of the lower protocol defines the local protocol
for \BuyerOne of the Book-store protocol. Role \BuyerOne
interacts with roles \BuyerTwo and \Seller. In essence,
the local protocol for role \BuyerOne is derived automatically
by stripping out all the irrelevant information with respect
to \BuyerOne from the global protocol.


%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Integrating Session Types
%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Integrating Session Types into Software Development}%: The Scribble Tool-chain}
\label{sec:sessions_integrate}

%The formal nature of session types allow for a rigorous manipulation
%of the communication behaviour inside a software.
\figref{diagram} illustrates a holistic approach to the integration of session
types in software engineering.
%, that follows procedures derived out of the rigorous
%nature of session types.
In \figref{diagram} rectangles
represent source code for either the software
under development (white coloured) or for the definition
of the session type protocols (grey coloured).
The arrows
represent automated procedures that ensure
the desirable properties of session types
in the development process. Blue arrows represent
a top-down view of the diagram while red arrows take
the bottom-up approach.
%The Scribble tool-chain supports many of these automated procedures.
An important assumption
is that the software source code in the
bottom of \figref{diagram} uses an interface
for communication, where the communication
operations can be identified by automated
procedures.
%
Furthermore and in addition to a protocol description language,
Scribble offers a tool-chain that implements most of the procedures
described in \figref{diagram}.

\input{figures/fig-sessions_software}

%Below the term system is used to describe the whole concurrent
%system and the term \dk{program} is used for communication module
%inside the system. \textbf{Bellow the term system is used to describe whole local protocol
%implementation as a set of communication modules it is composed of? }

%\dk{The formal nature of session types protocols
%	allow for rigorous manipulation of the communication behaviour.}

%\figref{diagram} clarifies the terminology
%developed previously by showing the
%\mybf{typical usage} of session types in software
%engineering.
%%design and development.
%Furthermore, it distinguishes
%the desirable properties that are offered by
%session types.


In a typical top-down approach, a developer will first develop a global protocol
%and express it syntactic form.
as syntax that expresses session type specifications.
This for example can be done
using the Scribble protocol description language.
The global protocol source code is shown
on the top of \figref{diagram}.
The next step is to run a procedure to verify the
global protocol's correctness: for example, checking that choices
are communicated to all roles whose subsequent behaviour depends on them.
After the global protocol verification, a projection
procedure takes place where a local protocol
for each global protocol role is derived.
The obtained local protocols are also expressed
syntactically as session type specifications.
Verification, projection and local specification
syntax are supported by the Scribble tool-chain.

\myparagraph{Type-check/Verification:}
One way to use the set of local protocols is to ensure
conformance between the source code of the
software under development and the local protocols,
by using a typechecking procedure.
%To achieve this there is the assumption that
%the programming language used to develop the
%software source code, makes use of a communication
%interface that can be identified by a type-checker
%procedure.
The typechecker takes as an input
%software module,
source code that corresponds to a communicating process;
it identifies its communication operations,
and derives the communication behaviour of
the
%module.
source code.
The typechecker also takes as an input the
corresponding syntax of the local protocol
and checks whether the
%software module
software source code
implements the corresponding role of the specified communication behaviour.
Upon success, the result is verified software that
enjoys the {\em communication safety} and {\em communication progress}
properties of session types.

\myparagraph{Code generation:}
A second way to use the set of local protocols
is for automatic code generation. The generated
code might be in the form of APIs that implement
communication behaviour, or in the form of program
templates that implement communication behaviour and
need manual work by the developers to add
computational logic.
Scribble supports the generation of a Java API for
each one of the set of Scribble local protocols.
The generated API has a class for each state of the protocol
that implements as methods all the available communication
operations. A method call returns an object of the class
that corresponds to the next state of the protocol. A dynamic
check ensures that each object (i.e.~state) is used exactly once.


\myparagraph{Monitoring:}
A third way of using the set of local protocols is
for monitoring the execution of the communication
interaction inside a concurrent system.
The local protocols are used to feed a set of local monitors.
Essentially, a local monitor runs a state machine that is derived
from the local protocol syntax.
%The monitor platform also requires for
%the concurrent software execution.
It is assumed that the communication interaction
of the software under execution can be recognised
and monitored by the monitor interface.
%The assumption to be made here is that
%monitor platform implements the communication interface
%on top of the monitoring interface. \textbf{I do not understand this sentence!}
Scribble supports the procedure of deriving
state machines from local protocol syntax.

\myparagraph{Inference/Synthesis:}
\figref{diagram} also shows a bottom-up approach,
where the communicating software source codes are developed independently.
Then an inference  procedure identifies the communication behaviour
of each component, resulting in a set of local protocols.
%\textbf{Here is not clear - if this is provided automatically?}.
After that, an automatic synthesis procedure can be used,
where the derived local protocols are synthesised
to create a global protocol. Successful synthesis of the global protocol guarantees that the
software has the communication safety and progress properties.


In any of the above methodologies, session types can be used as
requirements  to drive the software development process.
In the top-down approach, local types can be used to give intuition
and guide developers when implementing code.
In the bottom-up approach, more agile development teams
can analyse and compare the obtained global and/or local protocols
against the intended communication behaviour using human
reasoning techniques. Protocols can also be expressed
as diagram. For example, a local protocol is a syntactic form
of a state machine. Global protocols can be expressed as
petri-nets~\cite{DBLP:conf/tgc/FossatiHY14}, communicating
finite state machines~\cite{Denielou:2012:MST:2259248.2259258}
and BPMN choreographies~\cite{LTY2015}.





%\textbf{Furthermore, thus obtained global protocol can be compared
%to the specifications of intended communication behaviour,
%formally with tools or informally with human reading technique.}

%Typically, a protocol is first expressed globally and then
%projected locally \textbf{on each protocol role, projecting one local protocol specification per each protocol role.}
%Local specifications are then implemented by the corresponding
%\mybf{programs}.
%The term {\em session fidelity} is used to express
%the conformance of a \mybf{program}, to the local protocol of the
%corresponding role. Session fidelity can be checked mechanically,
%through techniques like type-checking and monitoring.

\myparagraph{Session fidelity:}
The case where the implementation of all the communication
processes inside a system conforms to the corresponding
local protocols of all the roles in a global protocol
%are implemented by the corresponding communicating modules of
is called {\em session fidelity}.
When session fidelity is ensured then the communication safety
and communication progress properties are ensured.
%then a number of communication properties can be guaranteed
%for the concurrent software:
%%
%\begin{enumerate}[label=$\bullet$]
%	\item	Every execution state of the system has {\em safe} communication behaviour:
%	%
%	\begin{itemize}
%		\item	{\em Communication interaction match}: Every send (resp., select) operation has a corresponding receive (resp., branch) operation.
%		\item	{\em Message (\mybf{Payload?}) type match}: The type of every message being send matches the type of the message expected to be receive.
%		\item	{\em Deadlock-freedom}: Consequently, every message send will be eventually received.
%	\end{itemize}
%
%	\item	Every state of the system conforms to a terminated protocol, or
%			it will {\em progress} to a state with a safe communication behaviour.
%\end{enumerate}

As seen in \figref{diagram} Scribble and the procedures
that manipulate session type protocols lie at the core
of the software development process. This is also clarified
by the experience of developing the usecases for the ABCD
online repository. The Scribble platform and its procedures
have an independent status next to the plethora of technologies and
programming paradigms used in the case studies.

%\subsection{Session Types as Protocols: The Scribble Tool-chain}
%
%This section presents the Scribble~\cite{scribble} tool-chain
%as the core tool for the application of session types.
%This section also uses Scribble to clarify the terminology
%developed in the previous section.

%The basic module of the Scribble tool-chain is the
%Scribble protocol description language,
%which is a syntax that expresses session type specifications.
%The design of the Scribble language draws directly from
%the principles of session types.

%\input{figures/fig-scribble_global_bookstore}

%In \figref{scribble_bs}, uses the Scribble protocol for the
%Bookstore usecase in the online repository~\cite{usecase_repository},
%to present the Scribble syntax.
%Line 1 defines the
%of global protocol \lstinline|Bookstore| between
%roles \lstinline|Buyer1|, \lstinline|Buyer2| and \lstinline|Seller|.
%%The communication structure is expressed
%%as a sequence of {\em endpoint} to endpoint message passing
%%interactions between the roles of the protocol.
%%Messages are described in terms of a message type,
%%where  a label is used to annotate a list
%%of types.
%Line 2 gives an example of the simplest endpoint-to-endpoint
%message passing interaction,
%where a message with
%label \lstinline|book| and type \lstinline|title| (notation \lstinline|book(title)|)
%is send from (keyword \lstinline|from|) role \lstinline|Buyer1| to (keyword \lstinline|to|)
%role \lstinline{Seller}.
%In Scribble messages are described as a type structure
%consisted of a label and a list of message types.


%Line 5 of the protocol clarifies the choice interaction, where
%role \lstinline|Buyer2| selects a choice from two possible outcomes,
%expressed with the labels \lstinline|agree| or \lstinline|quit|. In
%both cases roles \lstinline|Buyer1| and \lstinline|Seller| offer
%alternative branches.


%\input{figures/fig-scribble_local_bookstore}

%We use the term {\em local} protocol to describe
%the communication structure at the level of a single participant.
%As session types is a type discipline with the requirement to
%be enforced on a program, we may sometimes refer to a session type
%protocol as communication {\em specification}.

%For example Scribble code:
%%
%\begin{lstlisting}
%  msg(int) from A to B;
%\end{lstlisting}
%%
%describes the global protocol between role \lstinline|A| and \lstinline|B|,
%where participant \lstinline|A| sends message of type \lstinline|msg(int)|
%to participant \lstinline|B|. From the local perspective of participant
%\lstinline|A| the protocol would be:
%%
%\begin{lstlisting}
%  msg(int) to B;
%\end{lstlisting}
%%
%where it describes the sending of message \lstinline|msg(int)|
%to participant \lstinline|B|.


%\subsection{Fine/coarse grain representation}
%\dk{this should go as future work}

%{\em Concurrency patterns} e.g. network topology, client/server, race  condition, etc. (maybe I am writing nonsense here)



\subsection{Technologies based on Session Types}

We describe a selection of practical languages, frameworks and tools, based on the range of session types engineering methodologies described above, that we have used to implement the use-cases (describe in Section~\ref{sec:usecases}).
%%
%\begin{enumerate}
%	\item	Session Java~\cite{HU07TYPE-SAFE} is a Java extension
%			that statically checks binary session types for communication channels
%			that are implemented as an API.
%			%of Java with binary session types for communication channels, supported by a runtime library. The compiler statically checks session types.
%
%%	\item	Eventful Session Java~\cite{event} supports asynchronous, event-driven programming, using session types to track progress through individual sessions.
%
%	\item	Multiparty Session C~\cite{NYH12} supports programming in C with the MPI library. The compiler statically checks session types.
%
%	\item	SPY~\cite{DBLP:conf/rv/NeykovaYH13} uses session types for runtime monitoring of %communication
%			protocols in Python.
%
%	\item	Two technologies apply session types to the actors paradigm:
%			\begin{itemize}
%				\item	SPY is used to monitor Python threads that simulate actors~\cite{DBLP:conf/coordination/NeykovaY14}.
%%				\item	An implementation that uses SPY to
%%						monitor python threads that simulate actors~\cite{DBLP:conf/coordination/NeykovaY14}.
%				\item	Session type monitoring is also used for Erlang actors~\cite{SF15}.
%%				\item	A similar approach on monitoring session types is also used for the
%%						Erlang programming language. %, which is a programming language for the Actors model.
%%						\sg{Is this Erlang system one of ours? Simon Fowler's MSc}
%			\end{itemize}
%
%	\item	Pabble~\cite{DBLP:conf/pdp/NgY14} extends Scribble to express
%			communication structures that are parametrised by the number of roles.
%			Pabble protocols are implemented and typechecked for the C+MPI framework.
%
%	\item	Mungo~\cite{mungo} is a tool that integrates session types into the object-oriented
%%			programming
%			paradigm through the notion of typestate. Communication
%			operations on a channel are accessed via a state-dependent interface.
%			%Mungo uses a protocol description language for capturing typestate.
%			The Scribble to Mungo (StMungo) tool transforms Scribble
%			protocols into Mungo interfaces.
%
%
%	\item	GV:	A functional programming implementation of binary session types~\cite{DBLP:journals/jfp/Wadler14}.
%
%	\item	The web programming language LINKS~\cite{citation_needed}
%			uses the linear logic interpretation of session types~\cite{DBLP:journals/jfp/Wadler14}
%			to statically type-check protocols.
%
%	\item	The Scribble API generator, automatically creates an API for protocols,
%			where it describes each protocol state as a class equipped with
%			communication methods. Runtime checks ensure state linearity.
%
%%	\item	SILL: Functional programming based on the linear
%%			dual intuitionistic interpretation of
%%			session types. \sg{Separate this from the
%%			repository --- say something elsewhere.}
%
%\end{enumerate}


%
{\em Session Java}~\cite{HU07TYPE-SAFE}
a Java extension that statically checks binary session types for communication channels
that are implemented as an API.

{\em Multiparty Session C}~\cite{NYH12} supports programming in C with the MPI library. The compiler statically checks session types.

{\em SPY}~\cite{DBLP:conf/rv/NeykovaYH13} uses session types for runtime monitoring of %communication
protocols in Python.

Two technologies apply session types to the {\em actors} paradigm:
(i) SPY is used to monitor Python threads that simulate actors~\cite{DBLP:conf/coordination/NeykovaY14}; and
(ii)	Session type monitoring is also used for Erlang actors~\cite{SF15}.

{\em Pabble}~\cite{DBLP:conf/pdp/NgY14} extends Scribble to express
communication structures that are parametrised by the number of roles.
Pabble protocols are implemented and typechecked for the C+MPI framework.

{\em Mungo}~\cite{mungo} is a tool that integrates session types into the object-oriented
paradigm through the notion of typestate. Communication
operations on a channel are accessed via a state-dependent interface.
The Scribble to Mungo (StMungo) tool transforms Scribble
protocols into Mungo interfaces.


{\em GV} is functional programming implementation of binary session types~\cite{DBLP:journals/jfp/Wadler14}.

The web programming language {\em LINKS}~\cite{citation_needed}
uses the linear logic interpretation of session types~\cite{DBLP:journals/jfp/Wadler14}
to statically type-check protocols.

The {\em Scribble API generator}~\cite{citation_needed}, automatically creates an API for protocols,
where it describes each protocol state as a class equipped with
communication methods. Runtime checks ensure state linearity.

%	\item	SILL: Functional programming based on the linear
%			dual intuitionistic interpretation of
%			session types. \sg{Separate this from the
%			repository --- say something elsewhere.}





